How Slow do my Transient Typechecks Go?

### 2019-08-14 paper.tex
Spellchecked and compiles

\def\AWFY{Are\,We\,Fast\,Yet\xspace}

\usepackage{listings}

@@

There is a spectrum of different approaches to gradual typing
\cite{kafka18,bensurvey18icfp}.
At one end\,---\,``pluggable types'' as in Strongtalk \cite{strongtalk} or ``erasure
semantics'' as in 
%% Two important approaches are optional\citep{}
%% and gradual typing\citep{GiladPluggable2004,Siek2006,XXXSiek2015}.
%% These are applied to dynamic languages to reap the benefits of typing, but
%% unfortunately also have limitations.
%With optional or pluggable approaches such as 
%\citeurl{TypeScript,}{TypeScript}{Microsoft}{27 June
%2018}{https://www.typescriptlang.org/}
TypeScript\citep{typeScriptECOOP}\,---\,
%
all types are erased before the execution, limiting the benefit of
types to the statically typed parts of programs, and preventing
programs from depending on type checks at run time.  In the middle,
``transient'' or ``type-tag'' checks as in Reticulated Python 
% and ``concrete'' checks as in Thorn
offer first-order semantics, checking
whether an object's type constructor or supported methods match
explicit type declarations
%%%WRONG kjx
%, and checking only when data flows through those declarations
\cite{Siek2007,Bloom2009,concrete15,reticPython2014,Greenman2018}.
Reticulated Python also supports an alternative ``monotonic'' semantics
which mutates an object to narrow its concrete type when it is passed
into a more specific type context.
At the other end of the spectrum, behavioral
typechecks as in Typed Racket \cite{typedScheme08,takikawa2012},
Gradualtalk \cite{gradualtalk14},
and Reticulated Python's proxies,
support higher-order semantics, retaining
types until run 
time, performing the checks eagerly, and giving detailed information
about type violations as soon as possible via blame
tracking \cite{blame2009,blameForAll2011}.
Finally, Ductile typing
dynamically interprets a static type system at runtime
\cite{Ductile2011}.
%
%\mwh{Does a monotonic semantics fit on this spectrum somewhere?}
%\kjx{Not in the greenman/fellenstein spectrum, but I guess I'd put it
%  after concrete and before behavioural. Does it matter?}
%
Unfortunately, any gradual system with run-time semantics
(i.e.\ everything more complex than erasure) currently
imposes a significant run-time performance overhead to provide those
semantics.  This has lead to a significant body of research,
on one hand developing techniques to optimise gradual typing 
\citep{Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Greenman2018}
and on the other hand developing techniques to evaluate the
resulting performance  \cite{takikawa2016,Greenman2019jfp}.

@@

The static core of Grace's type system is well described
elsewhere\citep{TimJonesThesis};
here we explain how
these types can be understood 
dynamically, from the Grace programmer's point of view.
Grace's types are structural \citep{graceOnward12},
that is, an object implements a type whenever it
implements all the methods required by that type,
rather than requiring classes or objects to declare types explicitly.
Methods match when they have the same name and arity:
argument and return types are ignored.
A type thus expresses the requests an object can respond to,
for example whether a particular accessor is available,
rather than a nominal location in an explicit inheritance hierarchy.

@@

@@

In our previous work \cite{roberts-and-co-ecoop-2019} we evaluated the
performance of transient typechecks using the \AWFY
benchmark suite, plus other
benchmarks from the gradual-typing literature.
The goal was to complement our benchmarks with additional ones that are
used for similar experiments and can be ported to Grace.
To this end, we surveyed a number of papers\citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Stulova2016,Greenman2018}
and selected benchmarks that have been used by multiple papers.
Some of these benchmarks overlapped with the \AWFY suite,
or were available in different versions.
While not always behaviorally equivalent,
we chose the \AWFY versions since we already used them to
establish the performance baseline.
The selected benchmarks as well as the papers in which they were used are shown in
\cref{tab:gradual-benchmarks}.

@@

The benchmarks were modified to have complete type information.  To
ensure correctness and completeness of these experiments, we added an
additional check to Moth that reports absent type information to
ensure each benchmark is completely typed.  To assess the performance
overhead of type checking, we compared the execution of Moth with all
checks disabled against an execution that has all checks enabled.  The
results of this evaluation demonstrated that Moth run programs with
transient typechecks enabled at an overhead of of 5\% (min. −13\%,
max. 79\%) over peak performance without typechecks.

We did not measure programs that mix typed and untyped code because
with our implementation technique a fully typed program is expected to
have the largest overhead. This is not the case for other kinds of
gradual type checking, which have given rise to more complex
evaluation techniques.  In particular, so-called ``Takikawa'' or
``Takikawa-Greenman'' evaluation \cite{takikawa2016,Greenman2019jfp}
benchmarks a \emph{lattice} of $23^N$ \emph{configurations} of each
benchmark. where a configuration is a particular mix of static and
dynamically typed code.

The Takikawa evaluation protocol was originally proposed for Typed
Racket, where static vs dynamic typing is set per module, so $N$ is
the number of modules. Grace, following other languages such as
Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017},
allows programmers to choose typechecks at the level of individual
declarations: $N$ is the number of type annotations in the program.,
so checking an entire lattice for a benchmark of any size
is infeasible.  Vitousek et.al.\ modified the Takikawa protocol for
these kinds of langauges by taking an approach based on sampling
\cite{vitousek-transient-arXive-2019}.  The Takikawa-Vitousek protocol
divides the number of type annotations in an fully-typed program into
up to 100 intervals, and then randomly generates ten programs within
each interval by erasing type annotations.

@@

\textbf{
!!!KJX: Need to update this paragraph with what we did this time around!!!.
To account for the complex warmup behavior
of modern systems\citep{Barrett:2017:VMW} as well as
the non-determinism caused by \eg garbage collection and cache effects,
we run each benchmark for \NumIterationsAll iterations in the same
VM invocation.\footnote{
For the Higgs VM, we only use \NumIterationsHiggs iterations,
because of its lower performance.
This is sufficient since Higgs's compilation approach induces less variation
and leads to more stable measurements.}
Afterwards, we inspected the run-time plots over the iterations
and manually determined a cutoff of \WarmupCutOff iterations for warmup,
\ie, we discard iterations with signs of compilation.
As a result, we use a large number of data points to compute the average,
but outliers, caused by \eg garbage collection, remain visible in the plots.
% In this work, we do not consider startup performance,
% because we want to assess the impact of gradual type checks
% on the best possible performance.
All reported averages use the geometric mean since they aggregate ratios.

% Yuria
%  - Ubuntu 16.04.4, Kernel 3.13
%  - 24 hyperthreads
%  - Intel Xeon E5-2620 v3 2.40GHz
% Graal 0.33 Feb. 2018
%
% TODO: update these details when we have time... not a rejection criterion,
% but needs to be correct in the final version
All experiments were executed on a machine running Ubuntu Linux 16.04.4,
with Kernel 3.13.
The machine has two Intel Xeon E5-2620 v3 2.40GHz,
with 6 cores each, for a total of 24 hyperthreads.
We used ReBench 0.10.1\citep{ReBench:2018}, Java 1.8.0\_171, Graal 0.33 (\code{a13b888}),
Node.js 10.4, and Higgs from 9 May 2018 (\code{aa95240}).
Benchmarks were executed one by one to avoid interference between them.
The analysis of the results was done with R 3.4.1,
and plots are generated with ggplot 2.2.1 and tikzDevice 0.11.
Our experimental setup is available online to enable reproductions.\footnote{
  \smtodo{merge changes, and tag final version}
\url{https://github.com/gracelang/moth-benchmarks}}
}

@@

give overall resuslts - one page of monochromatic scatterplots one for
each benchmark.

@@

  - show that single annotation prediction doesn't reallty work\ldots

>>>>

\def\AWFY{Are\,We\,Fast\,Yet\xspace}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\ie}{i.e.\xspace}
\newcommand{\eg}{e.g.\xspace}

\usepackage{listings}
\usepackage{xspace}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{codered}{rgb}{0.82,0.15,0.23}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

@@

There is a spectrum of different approaches to gradual typing
\cite{kafka18,bensurvey18icfp}.
At one end\,---\,``pluggable types'' as in Strongtalk \cite{strongtalk} or ``erasure
semantics'' as in 
%% Two important approaches are optional\citep{}
%% and gradual typing\citep{GiladPluggable2004,Siek2006,XXXSiek2015}.
%% These are applied to dynamic languages to reap the benefits of typing, but
%% unfortunately also have limitations.
%With optional or pluggable approaches such as 
%\citeurl{TypeScript,}{TypeScript}{Microsoft}{27 June
%2018}{https://www.typescriptlang.org/}
TypeScript\citep{typeScriptECOOP}\,---\,
%
all types are erased before the execution, limiting the benefit of
types to the statically typed parts of programs, and preventing
programs from depending on type checks at run time.  In the middle,
``transient'' or ``type-tag'' checks as in Reticulated Python 
% and ``concrete'' checks as in Thorn
offer first-order semantics, checking
whether an object's type constructor or supported methods match
explicit type declarations
%%%WRONG kjx
%, and checking only when data flows through those declarations
\cite{Siek2007,Bloom2009,concrete15,reticPython2014,Greenman2018}.
Reticulated Python also supports an alternative ``monotonic'' semantics
which mutates an object to narrow its concrete type when it is passed
into a more specific type context.
At the other end of the spectrum, behavioral
typechecks as in Typed Racket \cite{typedScheme08,takikawa2012},
Gradualtalk \cite{gradualtalk14},
and Reticulated Python's proxies,
support higher-order semantics, retaining
types until run 
time, performing the checks eagerly, and giving detailed information
about type violations as soon as possible via blame
tracking \cite{blame2009,blameForAll2011}.
Finally, Ductile typing
dynamically interprets a static type system at runtime
\cite{Ductile2011}.
%
%\mwh{Does a monotonic semantics fit on this spectrum somewhere?}
%\kjx{Not in the greenman/fellenstein spectrum, but I guess I'd put it
%  after concrete and before behavioural. Does it matter?}
%
Unfortunately, any gradual system with run-time semantics
(i.e.\ everything more complex than erasure) currently
imposes a significant run-time performance overhead to provide those
semantics.  This has lead to a significant body of research,
on one hand developing techniques to optimise gradual typing 
\citep{Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Greenman2018}
and on the other hand developing techniques to evaluate the
resulting performance  \cite{Takikawa2016,Greenman2019jfp}.

@@

The static core of Grace's type system is well described
elsewhere\citep{TimJonesThesis};
here we explain how
these types can be understood 
dynamically, from the Grace programmer's point of view.
Grace's types are structural \citep{graceOnward12},
that is, an object implements a type whenever it
implements all the methods required by that type,
rather than requiring classes or objects to declare types explicitly.
Grace's transient dynamic typechecks are first order, so methods match when they have
the same name and arity: 
argument and return types are ignored.
A type thus expresses the requests an object can respond to,
for example whether a particular accessor is available,
rather than a nominal location in an explicit inheritance hierarchy.

@@

To see how this works in practice, consider this code snippet, which
will run in Grace.

\begin{lstlisting}
def o = object {
   method three -> Number {3} }
type ThreeString = interface {
   three -> String
}
def t : ThreeString = o
printNumber (t.three)
\end{lstlisting}

Dynamic typechecks will take place: on line 6, when object \code{o}
initialises the variable \code{t}, we check it has the method
\code{three}; on line 7, when the value of \code{t} is read, it must
again have the method \code{three}; on line 2, when the \code{three}
method requested by ``\code{t.three}'' returns, the return value must
be a \code{Number}; and (presumably) within the definition of 
%
\code{printNumber(n :   Number)} 
%
(not shown) to check that its argument is a number. Note that we never check
whether e.g.\ the result of requesting ``\code{t.three}'' is actually
a \code{String} (as one may expect from line 4) because we check
only first-order types (whether objects have conforming methods) not higher-order
types (whether those methods' argument and result types conform), and
we check only when values flow through explicit type annotations.
This is why the type declared in lines 3-5 is checked only on line 6
(where it is mentioned explicitly); and the check only requires the
presence of a method called \code{three}, regardless of the method's
declared return type.

@@

In our previous work \cite{roberts-and-co-ecoop-2019} we evaluated the
performance of transient typechecks using the \AWFY
benchmark suite, plus other
benchmarks from the gradual-typing literature.
The goal was to complement our benchmarks with additional ones that are
used for similar experiments and can be ported to Grace.
To this end, we surveyed a number of papers\citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Stulova2016,Greenman2018}
and selected benchmarks that have been used by multiple papers.
Some of these benchmarks overlapped with the \AWFY suite,
or were available in different versions.
While not always behaviorally equivalent,
we chose the \AWFY versions since we already used them to
establish the performance baseline.
The selected benchmarks as well as the papers in which they were used are shown in
Table~\ref{tab:gradual-benchmarks}.

@@

The benchmarks were modified to have complete type information.  To
ensure correctness and completeness of these experiments, we added an
additional check to Moth that reports absent type information to
ensure each benchmark is completely typed.  To assess the performance
overhead of type checking, we compared the execution of Moth with all
checks disabled against an execution that has all checks enabled.  The
results of this evaluation demonstrated that Moth run programs with
transient typechecks enabled at an overhead of of 5\% (min.\ -13\%,
max. 79\%) over peak performance without typechecks.

We did not measure programs that mix typed and untyped code because
with our implementation technique a fully typed program is expected to
have the largest overhead. This is not the case for other kinds of
gradual type checking, which have given rise to more complex
evaluation techniques.  In particular, so-called ``Takikawa'' or
``Takikawa-Greenman'' evaluation \cite{Takikawa2016,Greenman2019jfp}
benchmarks a \emph{lattice} of $23^N$ \emph{configurations} of each
benchmark. where a configuration is a particular mix of static and
dynamically typed code.

The Takikawa evaluation protocol was originally proposed for Typed
Racket, where static vs dynamic typing is set per module, so $N$ is
the number of modules. Grace, following other languages such as
Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017},
allows programmers to choose typechecks at the level of individual
declarations: $N$ is the number of type annotations in the program.,
so checking an entire lattice for a benchmark of any size
is infeasible.  Vitousek et.al.\ modified the Takikawa protocol for
these kinds of languages by taking an approach based on sampling
\cite{vitousek-transient-arXive-2019}.  The Takikawa-Vitousek protocol
divides the number of type annotations in an fully-typed program into
up to 100 intervals, and then randomly generates ten programs within
each interval by erasing type annotations.

@@

\textbf{!!!KJX: Need to update this paragraph with what we did this time around!!!}
To account for the complex warmup behavior
of modern systems\citep{Barrett:2017:VMW} as well as
the non-determinism caused by \eg garbage collection and cache effects,
we run each benchmark for \textbf{$/$NumIterationsAll} iterations in the same
VM invocation.\footnote{
For the Higgs VM, we only use \textbf{$/$NumIterationsHiggs} iterations,
because of its lower performance.
This is sufficient since Higgs's compilation approach induces less variation
and leads to more stable measurements.}
Afterwards, we inspected the run-time plots over the iterations
and manually determined a cutoff of \textbf{$/$WarmupCutOff} iterations for warmup,
\ie, we discard iterations with signs of compilation.
As a result, we use a large number of data points to compute the average,
but outliers, caused by \eg garbage collection, remain visible in the plots.
% In this work, we do not consider startup performance,
% because we want to assess the impact of gradual type checks
% on the best possible performance.
All reported averages use the geometric mean since they aggregate
ratios.

\textbf{
% Yuria
%  - Ubuntu 16.04.4, Kernel 3.13
%  - 24 hyperthreads
%  - Intel Xeon E5-2620 v3 2.40GHz
% Graal 0.33 Feb. 2018
%
% TODO: update these details when we have time... not a rejection criterion,
% but needs to be correct in the final version
All experiments were executed on a machine running Ubuntu Linux 16.04.4,
with Kernel 3.13.
The machine has two Intel Xeon E5-2620 v3 2.40GHz,
with 6 cores each, for a total of 24 hyperthreads.
We used ReBench 0.10.1\citep{ReBench:2018}, Java 1.8.0\_171, Graal 0.33 (\code{a13b888}),
Node.js 10.4, and Higgs from 9 May 2018 (\code{aa95240}).
Benchmarks were executed one by one to avoid interference between them.
The analysis of the results was done with R 3.4.1,
and plots are generated with ggplot 2.2.1 and tikzDevice 0.11.
Our experimental setup is available online to enable reproductions.\footnote{
\url{https://github.com/gracelang/moth-benchmarks}}
}

@@

give overall results - one page of monochromatic scatterplots one for
each benchmark.

@@

  - show that single annotation prediction doesn't really work\ldots
